---
title: "Lab 01: Introduction to R"
description: >
  Learn the basics of R, how to comment, and about the major pieces we'll be using in PSY 203.
author: Justin Dainer-Best
tutorial:
  id: "PSY-203-F20"
  version: 0.2
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: lumen
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(gradethis)
library(flair)

tutorial_options(exercise.timelimit = 200, exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(error = TRUE)
```

## Introduction

We'll be using this sort of web document in conjunction with R Markdown documents during the course of this class. 

Much of the data we'll use in learning is what I would call "tame" data---cleaned up, easy to understand, and without complications. But we'll learn to process "messy" (real-world) data, too. 

### Why R?

[Watch me discuss these points in a video on Brightspace](https://bardcollege.brightspace.com/d2l/le/content/6809/viewContent/11807/)

1. Free & Open Source (not proprietary)
2. Flexible
3. Widely used across disciplines (including outside of the Bard bubble!)
4. Introduction to code and to programming
5. Really fun!
6. Tons of free online resources (and print books, if you prefer)
7. Excellent visualization ability
8. Integrating theory and practice
9. R Notebooks and Markdown files are great ways to follow along!
10. Packages and extensibility

### Script vs. R Markdown vs. Console

1. You can write code in an R Markdown or R Notebook document, which combines code with text.
2. You can also write in a "script" window (in R or R Markdown) where all text is treated as code.
3. You can just write code directly into the console.
4. You can also write code in a text editor, and then paste it in. 

I will usually use R Markdown in work that we do together---but all of these options make sense! 

You have RStudio open on your computer (or through RStudio.Cloud). You might want to take notes *in RStudio* as you complete these exercises. You can alternate between windows and open an R Notebook with File: New File: R Notebook. You won't be turning this in today; this is just for you.

## Starting to play around with R

R does *not* always need you to enter all of your code on one line. Try running the next few lines all at once by clicking "Run code" at the top right. 

Did you get your expected result for each addition? (Each answer will be in order, below.)

```{r basicmath, exercise=TRUE}
3 + 4

3 +
  4

8 + 4

8 
+ 4
```

<div id="basicmath-hint">
**Hint:** Try putting 8 + 4 on the same line.
</div>

In essence, R "waits" for you to finish if and only if it is very obvious that you aren't done. White space (spaces, tabs, etc.) doesn't matter *unless* the line is obviously done---on the second-to-last line, R reads 8---a perfectly complete statement---and executes it. (Similarly, +4 is a positive 4.) On the other hand, 3 + is *not* a complete statement---so R waits for you to continue it.

What are some examples of characters that make it clear that the line isn't done? Try playing around in the above area, or do it in your own R console. (Some suggestions to test---which does which?: parentheses, mathematical operators, decimals.)

### Exercise

Use the below space to write R code to find the sum of 10 and 4, to subtract 8 from 10, and to multiply 3 by 7 (use the asterisk * for multiplication). Click submit answer (at the top right) when you're done.

```{r add_subtract, exercise=TRUE, exercise.lines = 3}

```

```{r add_subtract-solution}
10 + 4
10 - 8
3 * 7
```

<div id="add_subtract-hint">
**Hint:** Make sure each of the operations is on a separate line. Additionally, the order matters. (e.g., I wrote "subtract 8 from 10", not "10 from 8")
</div>

```{r add_subtract-check}
grade_code("Nicely done!")
```

## Commenting

Any text written in the R Markdown document appears as regular text. However, when writing code, we frequently also write *comments* to remind us what happens. Those are done using a # in R code: 

```{r comments, include=FALSE}
# This is a comment
# ... which is within an R Markdown document

# Comments above code: this is addition
5 + 5

exp(3) # This is code at the end of a line

help(exp)
# This comment follows the code -- I recommend against it (but you can do this if you choose)
# help() can be very useful!
# ?exp would do the same thing.
```

```{r comments-2, echo=FALSE}
decorate("comments") %>% 
  flair("#")
```

### Exercise

Write R code to divide 144 by 12 (using the / for divide). Then **on the same line** add a comment that explains what you're doing. 

```{r comment, exercise=TRUE, exercise.lines = 1}

```

```{r comment-solution}
144/12 # this divides 144 by 12
```


## Variables and classes

Any type of coding relies on **variables**---words/letters that represent values that are likely to change. This is the programming equivalent of "*x*" in an equation like $y=x^2$---the *x* can take on different values, right? The same is true in R. And in fact, we'll frequently use variables like *x* in R, too. The point of a variable is that it can be re-used. The example we'll use for a moment is just a number---but we can store much larger data in a single variable.

*x* is a _variable_ that can hold a value. The most basic version is assigning it to a number like 5. We **assign** using what is called the assignment operation -- `<-` (a less than sign followed by a dash) -- think of that symbol as meaning "assign to". You assign the **value** on the right to the **variable** on the left. This is probably the most important thing in today's lesson! 

* Also note: when you see text `that looks like this`, that should usually mean that it's code.

(You can read more about variables in R [on this website](https://intro-r.rsquaredacademy.com/variables.html).)

You can also use an equal sign (=), but you'll see that when we are *creating a variable*, I will always use the assignment operator, like this:

```{r definex, include=FALSE}
x <- 5
x = 5
```

```{r definex-flair, echo=FALSE}
decorate("definex") %>% 
  flair("<-") %>%
  flair("=")
```

In this case, the code has already run. Read it aloud as: "I assigned the value of 5 to the variable x" and "I set x to be equal to 5." 

Try it yourself by running the following code:

```{r definex_try, exercise=TRUE}
x <- 5
```

You'll note that R doesn't print anything out when we assign this value. That's because you're not telling it to print anything---you're just recording a value. If you copy the code `x <- 5` into your R console and run it there, you'll see that it adds a variables called *x* into your global environment (at the rop right of RStudio). Its value should be 5. 

To print it out, you can just write the *name* of the variable---*x*. Try that, below:

```{r printx, exercise=TRUE, exercise.setup = "definex_try"}

```

```{r printx-solution}
x
```

```{r printx-check}
grade_code("Nicely done!")
```

<div id="printx-hint">
**Hint:** You just need to put the name of the variable in the exercise area.
</div>

### Classes (types) of variables

Each variable is a **class** or **type** of data. We'll learn about types of data in class, too---they're slightly different in R, because they're based on what data a computer can understand. 

In class, we'll talk about the following kinds of data: nominal (categorical), rank order (ordinal), interval (numeric), ratio (interval with absolute zero). We'll also define data as discrete or continuous. In R, there are more data **classes**, because each class refers to how data is stores. 

```{r datatypes, echo=FALSE}
tibble::tribble(
  ~`Data Type`, ~`R Class`,
  "nominal", "factor or character",
  "ordinal", "factor or numeric",
  "interval", "numeric", 
  "ratio", "numeric"
) %>% gt::gt()
```

We can use the word class followed by an opening parenthesis, the name of the variable, and a closing parenthesis to find out the class (or type) of variable x is---try running this code:

```{r class_of_x, exercise=TRUE}
class(x)
```

### More than one value

Now, let's make this just a bit more complex: a variable can represent more than one number at a time. You can store a **combination** of numbers in a variable by using the combine command or `c()`. Let's try it with another variable still called *x*---run the following code:

```{r c_of_x, exercise=TRUE, exercise.setup = "definex_try"}
x <- c(4, 5, 7)
```

Now, you can again just write the *name* of the variable---*x*. Try that, below:

```{r printx-2, exercise=TRUE, exercise.setup = "c_of_x"}

```

```{r printx-2-check}
grade_result(
  pass_if(~ identical(.result, c(4, 5, 7)), "Yes, you assigned c(4, 5, 7) to the variable x, so this is the expected output."),
  fail_if(~ identical(.result, 4), "Did you assign x to the value immediately above? Make sure to run the code, and feel free to click the 'start over' in the box if you can't."),
  fail_if(~ identical(.result, 5), "Did you assign x to the value immediately above? Make sure to run that code."),
  fail_if(~ TRUE, "Not quite. All that should be in the window is the variable name -- and that variable name should be assigned to the values of 4, 5, and 7 as above.")
)
```

And again use the `class()` function to determine the type of variable that *x* is:

```{r class_x, exercise=TRUE, exercise.setup = "c_of_x"}

```

```{r class_x-solution}
class(x)
```

```{r class_x-check}
grade_code("Nicely done!")
```

Why is the class the same? These are bout *vectors* of numbers---matrices with one row/column. Feel free to look up the terms *vector* and *matrix*, but I think the major point here is that a series of numbers are still numeric! 

## Data frames

Another useful type of variable is a structure called a *data frame*. It is essentially a two-dimensional matrix of data---like the sort of spreadsheet you see in Excel. In R, the data frame has one name (that's the main variable) and then all of the columns also have names (those are variables that exist within the data frame). 

Think of it like this: you've collected data on experiences of stigma on a college campus. This is your data set, and you store it in a data frame called **stigma**. That's the main variable. Within the data frame are columns of the individual variables you collected, things like where the experience occurred (**where**), the age of the person experiencing stigma (**age**), and what kind of interaction there was (**interaction_type**). Imagine it looks something like this:

```{r, echo=FALSE}
tibble::tribble(
  ~where, ~age, ~interaction_type,
  "library", 19, "in-person",
  "campus road", 20, "online"
)
```
The column headers in a data frame are those variable names. You'll see below the name the `class()` of each variable, too. 

So how would we create a variable like this? There are a couple of ways---this is one of the nice/frustrating things about R! The most basic command is `data.frame()` -- let's use that one now:

```{r df1, exercise = TRUE}
data.frame(
  age = c(18, 19, 20)
  )
```

So what does that do? As you may remember, the extra space doesn't mean anything---it just makes it easier to read. That's because the open parenthesis means R *expects* a closing parenthesis.

Read it aloud: "Create a data frame with a single column variable called age, and make age equal to the combination of 18, 19, and 20." 

Note that we haven't "named" this new data frame---that's why it prints to the screen. Within the data frame, as we're naming the variables, we **don't** use the assignment operator (<-). That operator is only for the sort of "top-tier" assignment. You'll note that if you copy and paste the above code (repeated below) into the RStudio console and run it, nothing shows up in the "Environment" pane at the top right. Again, that's because we haven't created a variable as R understands it. 

So, give this a try: at the beginning of the code below, create a variable *named `ages`* which is assigned (using `<-`) to the data.frame() code I've written for you:

```{r df2, exercise = TRUE, exercise.warn_invisible = FALSE}
data.frame(
  age = c(18, 19, 20)
  )
```

```{r df2-solution}
ages <- data.frame(
  age = c(18, 19, 20)
  )
```

```{r df2-check}
grade_code("Nicely done!")
```

If we want to add a second column in that data frame, we just add a comma at the end of the `age` line, and hit enter:

```{r df2_part2, exercise = TRUE}
data.frame(
  age = c(18, 19, 20),
  gender = c("f", "nb", "f")
  )
```

The quotation marks tell R that the "f" isn't a variable---it's character (text) input. 

### Exercise

Okay, try it yourself. Do the following below:

1. Create a variable called `votes`
2. Its value will be a `data.frame()`
3. Containing the following two variables:
    a. `party` equal to the **combined** (`c()`) values of `"Democrat", "Republican", "Independent"`
    b. `vote` equal to the **combined** (`c()`) values of `"Incumbent", "Challenger", "Incumbent"`

Some important things to remember:

1. Comment your code!
2. Close parentheses (but also note: RStudio will help you do that! So make sure there are the right number.)
3. Spell things correctly, and don't capitalize except in text commands.
4. The assignment operator `<-` is used in the initial variable name, but NOT within the `data.frame()` command.
5. You can absolutely copy and paste from above.
6. You'll need a comma between each character input (as I inserted them above) **and also** between the `c()` commands---as above
7. If you're not getting the "right" answer, that's okay! You can click the Solution button---what was wrong? The checker is just trying to see if you got the same code I wrote---minor errors aren't a big deal, and will always get full points.

```{r df3, exercise = TRUE, exercise.warn_invisible = FALSE, exercise.lines = 5}

```

```{r df3-solution}
votes <- data.frame(
  party = c("Democrat", "Republican", "Independent"),
  vote = c("Incumbent", "Challenger", "Incumbent")
)
```

```{r df3-check}
grade_code("Nicely done!")
```

When you write code like this, you'll see that R highlights problems---like an un-closed parenthesis. If you make any mistakes, you'll also see that the errors it gives might be a bit vague... you can always ask me when you have questions! 

## Review

Here are a few quiz questions to make sure you're following along. 

```{r quiz}
quiz(
  question("When do we use the assignment operator `<-`?",
    answer("When assigning the names of columns in a data frame"),
    answer("To print the value of a variable"),
    answer("When assigning a top-level variable, like the name of a data frame", correct = TRUE),
    answer("Never")
  ),
  question("Which of the following are R data types (classes)?",
    answer("numeric", correct = TRUE),
    answer("interval"),
    answer("categorical"),
    answer("character", correct = TRUE)
  )
)
```


## Subsets of data

